---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# PTSDdiag

<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version/PTSDdiag)](https://CRAN.R-project.org/package=PTSDdiag)
[![R-CMD-check](https://github.com/TobiasRSpiller/PTSDdiag/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/TobiasRSpiller/PTSDdiag/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

<p>
<a href="https://tobiasrspiller.github.io/PTSDdiag/articles/quickstart.html" class="btn btn-primary btn-lg">Quick Start</a>
<a href="https://tobiasrspiller.github.io/PTSDdiag/articles/internal_analysis.html" class="btn btn-outline-primary btn-lg">Full Analysis</a>
<a href="https://tobiasrspiller.github.io/PTSDdiag/reference/index.html" class="btn btn-outline-primary btn-lg">Reference</a>
<a href="https://doi.org/10.31219/osf.io/6rk72_v1" class="btn btn-outline-secondary btn-lg">Publication</a>
</p>

## Description 
PTSDdiag is a comprehensive R package for analyzing and simplifying PTSD diagnostic criteria using PCL-5 (PTSD Checklist for DSM-5) data. It provides tools to identify optimal subsets of six PCL-5 items that maintain diagnostic accuracy while reducing assessment burden. 

## **Key Features**

 * Data preparation and standardization for PCL-5 scores
 * Implementation of DSM-5 diagnostic criteria
 * Calculation of diagnostic metrics and summary statistics
 * Simplification of diagnostic criteria through:
    * Hierarchical (cluster-based) approach
    * Non-hierarchical approach
 * Comparison of different diagnostic approaches
 * Model validation using:
    * Holdout Validation
    * Cross-Validation

## Installation

You can install the released version of PTSDdiag from [CRAN](https://CRAN.R-project.org/package=PTSDdiag):

``` {r, eval = FALSE}
install.packages("PTSDdiag")
```

Or install the development version from GitHub:

``` {r, eval = FALSE}
# install.packages("devtools")
devtools::install_github("TobiasRSpiller/PTSDdiag")
```

## Getting Started

The vignette demonstrates how to use the package to prepare the PCL-5 data, calculate some basic descriptive statistics and reliability metrics, find the optimal minimal symptom combinations for PTSD diagnosis, compare different diagnostic approaches and perform validation methods for evaluating model performance.

 * [Quick Start](https://tobiasrspiller.github.io/PTSDdiag/articles/quickstart.html)
 * [Full Internal Analysis](https://tobiasrspiller.github.io/PTSDdiag/articles/internal_analysis.html)
 * [External Validation (PCL-5)](https://tobiasrspiller.github.io/PTSDdiag/articles/external_validation_pcl5.html)
 * [External Validation (CAPS-5)](https://tobiasrspiller.github.io/PTSDdiag/articles/external_validation_caps5.html)
 * [Package Website](https://tobiasrspiller.github.io/PTSDdiag/)
 
## Bugs, Contributions

* If you have any suggestions or if you find a bug, please report them using GitHub [issue tracker](https://github.com/TobiasRSpiller/PTSDdiag/issues).
* Contributions are welcome! Please feel free to submit a Pull Request.
