---
title: "CAPS-5 workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{CAPS-5 workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4
)
```

This vignette compares a PCL-5 symptom definition against the CAPS-5 diagnosis, and shows how to optimize a definition against the CAPS-5 diagnosis directly.

## Why use the CAPS-5

The PCL-5 is completed by the patient; the CAPS-5 is administered by a clinician. Both rate the same 20 DSM-5-TR symptoms on the same 0 to 4 scale, so when both are available for the same participants the CAPS-5 diagnosis is the closest available reference to a clinical ground truth. That makes it the natural standard against which to judge a self-report definition.

## Requirements for the input data

The CAPS-5 input must be the 20 items in their standard order, scored 0 to 4, with no missing values, plus any identifier columns you name in `id_col`; the full contract is in the [Getting started](getting-started.html) vignette. One **additional requirement** applies here: when comparing two instruments, the PCL-5 and CAPS-5 data frames must describe the same participants in the same row order.

## Standardizing both instruments

We standardize the PCL-5 data as usual and create a CAPS-5 data frame in the same form with `rename_caps5_columns()`. The CAPS-5 ratings below are simulated independently of the PCL-5 scores, purely for illustration. Because they are random rather than clinician-rated, the agreement reported in the next section reflects chance, and the tables should be read as illustrations of the output format rather than as a real instrument comparison. We use a 500-row subset of the bundled data to keep the vignette fast.

```{r setup, message = FALSE}
library(PTSDdiag)

data("simulated_ptsd")
pcl5 <- simulated_ptsd[1:500, ]
ptsd <- rename_ptsd_columns(pcl5,
                            id_col = c("patient_id", "age", "sex"))

set.seed(42)
caps5_raw <- as.data.frame(matrix(
  sample(0:4, 20 * nrow(pcl5), replace = TRUE),
  ncol = 20
))
caps5 <- rename_caps5_columns(caps5_raw)
```

## Computing the CAPS-5 diagnosis

`create_caps5_diagnosis()` applies the DSM-5-TR algorithm to the CAPS-5 severity scores, using the same clusters and the same presence threshold of 2 or higher as the PCL-5 diagnosis. It returns a single logical column, `PTSD_caps5`.

```{r caps5-dx}
caps5_dx <- create_caps5_diagnosis(caps5)
head(caps5_dx)
mean(caps5_dx$PTSD_caps5) * 100
```

## Comparing the PCL-5 and CAPS-5 diagnoses

`compare_diagnostic_systems()` builds one summary table that scores each diagnostic system against a chosen reference. Setting `reference = "caps5"` makes the CAPS-5 the standard, so the PCL-5 diagnosis and ICD-11 are evaluated against it. Each row reports sensitivity, specificity, PPV, and NPV against the reference, together with the counts of false positives, false negatives, and total misclassifications. 

```{r compare}
compare_diagnostic_systems(
  ptsd,
  caps5_data = caps5,
  icd11      = TRUE,
  reference  = "caps5"
)
```

## Optimizing against the CAPS-5 diagnosis

To make the CAPS-5 the optimization target rather than a row in a comparison, pass the clinician diagnosis to `compare_optimizations()` as a fixed criterion. The optimized PCL-5 definitions are then reported alongside it, and the symptom-frequency tools apply as in the [Comparing diagnostic criteria](comparing-criteria.html) vignette.

```{r caps5-as-fixed}
comp <- compare_optimizations(
  ptsd,
  scenarios = list(
    "4/6 Hierarchical"     = list(n_symptoms = 6, n_required = 4, hierarchical = TRUE),
    "4/6 Non-hierarchical" = list(n_symptoms = 6, n_required = 4, hierarchical = FALSE),
    "CAPS-5 (reference)"   = list(type = "fixed", criterion = caps5_dx$PTSD_caps5,
                                  symptoms = 1:20)
  ),
  n_top         = 5,
  score_by      = "accuracy",
  show_progress = FALSE
)
summarize_top_combinations(comp, top_n = 3, as_percent = TRUE)
```

## Interpreting agreement and disagreement

When the PCL-5 and CAPS-5 diagnoses agree closely, a simplified PCL-5 definition that reproduces the full PCL-5 diagnosis also reproduces the clinician diagnosis, and the choice of reference is immaterial. When they diverge, the choice of reference matters, and reporting performance against both is desirable.

## See also

- [Getting started](getting-started.html) for the single-cohort derivation workflow.
- [Comparing diagnostic criteria](comparing-criteria.html) for multi-rule comparison and the symptom-frequency heatmap.
- [Validating abbreviated symptom definitions](validation.html) for internal and external validation.
