---
title: "Getting started with PTSDdiag"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with PTSDdiag}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4
)
```

PTSDdiag identifies simplified PTSD diagnostic criteria that maintain classification accuracy against the full DSM-5-TR criteria.

## The problem addressed by PTSDdiag

PTSD as defined by the DSM-5-TR requires the presence of symptoms across four clusters, drawing from a total of 20 symptoms. This polythetic, hierarchical structure theoretically permits over 636,000 symptom combinations qualifying for a diagnosis, yet empirical research has shown that patients cluster around a limited number of distinct profiles. The complexity of the full criteria creates a substantial assessment burden and contributes to diagnostic heterogeneity that complicates both clinical practice and research. The PTSDdiag package implements a computational optimization approach that systematically searches all possible subsets of a given size, identifies the combinations that minimize misclassification relative to the full DSM-5-TR criteria, and evaluates the robustness of the results across multiple high-performing solutions. This vignette demonstrates a complete workflow from a scored PCL-5 dataset to optimized symptom combinations and summary output.

## Requirements for the input data

The optimization operates on the 20 PCL-5 items, so the package first needs to know which columns hold those items and which columns identify the participant. `rename_ptsd_columns()` does both: it renames the item columns to a canonical `symptom_1` through `symptom_20`, and it leaves any identifier columns you name in `id_col` untouched beside them.

Four caveats are worth mentioning.

- **Items are matched by position, not by name.** The first non-identifier column becomes `symptom_1`, the second becomes `symptom_2`, and so on. The item columns must therefore already be in the standard DSM-5-TR order (intrusive memories first, sleep disturbance last).
- **The data may contain only the 20 items and the columns you name in `id_col`.** Because matching is positional, the function cannot distinguish a stray column (an item from another questionnaire, a date) from a PCL-5 item. Rather than risk shifting every item by one position, it stops with an error whenever the number of non-identifier columns is not exactly 20. List the columns you want to keep (e.g., identifiers, demographics) in `id_col`, and remove anything else before calling.
- **Items scoring needs to follow the PCL-5 convention.** Items need to be scored 0 to 4, and a symptom counts as present at a score of 2 or higher, following the DSM-5-TR. 
- **Missing data is not permitted.** Missing data is handled explicitly by stopping the function because it could otherwise bias the results.

## Importing and preparing the data

We use the bundled `simulated_ptsd` dataset, which holds 5,000 simulated records with three demographic columns (`patient_id`, `age`, `sex`) followed by the 20 PCL-5 items (`S1`–`S20`).

```{r setup, message = FALSE}
library(PTSDdiag)
library(dplyr)

data("simulated_ptsd")
dim(simulated_ptsd)
names(simulated_ptsd)[1:6]
```

We pass the three demographic columns to `id_col`. They stay attached to every participant through the rest of the workflow, which is what later lets us describe the sample without a separate merge.

```{r rename}
ptsd <- rename_ptsd_columns(simulated_ptsd,
                            id_col = c("patient_id", "age", "sex"))
names(ptsd)[1:6]
```

The item columns are now `symptom_1` (intrusive memories) through `symptom_20` (sleep disturbance), with the demographics preserved in front.

## Defining the reference diagnosis

Every simplified definition is then judged against the full DSM-5-TR diagnosis, so we need to compute the "true" diagnosis first. `create_ptsd_diagnosis_nonbinarized()` applies the DSM-5-TR algorithm to all 20 items and adds a logical column, `PTSD_orig`, which is `TRUE` for participants who meet the criteria and `FALSE` otherwise. This column is the outcome that the optimization will try to reproduce with fewer symptoms.

A diagnosis requires at least one symptom from the intrusion cluster (items 1–5), at least one from avoidance (items 6–7), at least two from negative alterations in cognition and mood (items 8–14), and at least two from alterations in arousal and reactivity (items 15–20), each symptom counted as present at a score of 2 or higher.

```{r diagnosis}
ptsd <- ptsd %>%
  create_ptsd_diagnosis_nonbinarized() %>%
  calculate_ptsd_total()
```

`calculate_ptsd_total()` adds the PCL-5 sum score (range 0–80), a severity index we will use to describe the sample. In this clinical-style dataset, `r round(mean(ptsd$PTSD_orig) * 100)`% of participants meet the full criteria.

```{r prevalence}
mean(ptsd$PTSD_orig) * 100
```

Here is the rewritten section. All prose now sits before the code, the example is framed as a four-of-six definition that ignores the cluster structure, the combination count is given explicitly, the `score_by` choice is explained with both options and a rationale, and `n_top` and `show_progress` are covered. No dashes used.

---

## Identifying a symptom subset

With the reference diagnosis in place, `optimize_combinations()` searches for the symptom subset that reproduces it most closely. In this example we test a simplified definition that requires four of the six selected symptoms to be present, without taking the DSM-5-TR cluster structure into account.

For a subset of six items, the function evaluates every possible combination of the 20 symptoms. The number of combinations is $\binom{20}{6} = 38{,}760$. To each one it applies the presence rule, classifying a participant as positive when at least `n_required` of the six selected symptoms are present (scored 2 or higher), and then ranks the combinations by how often their decision disagrees with `PTSD_orig`.

Before running the search we choose the metric to optimize through the `score_by` argument. Two options are available. Setting it to `"accuracy"` minimizes the total number of misclassifications, counting false positives and false negatives equally. Setting it to `"sensitivity"` minimizes false negatives only, which prioritizes not missing participants who meet the full criteria, at the cost of more false positives. In this example we optimize for accuracy, because our aim is a simplified definition that agrees with the full DSM-5-TR diagnosis overall rather than one tuned toward a single type of error.

Two further arguments shape the output. `n_top = 3` returns the three best scoring subsets rather than a single winner, so that near optimal solutions can be compared; the companion value `res$n_tied` reports how many other combinations matched the top score exactly, which shows whether the leading subset is uniquely best or one of several equivalent solutions. `show_progress = FALSE` suppresses the progress bar.

```{r optimize, message = FALSE}
res <- optimize_combinations(
  ptsd,
  n_symptoms    = 6,
  n_required    = 4,
  n_top         = 3,
  score_by      = "accuracy",
  show_progress = FALSE
)
res$best_symptoms
```

## Diagnostic performance metrics

The optimization minimizes the total number of misclassified participants, but a single error count hides the kind of error being made, and the two kinds carry different costs. A false negative is a participant who meets the full DSM-5-TR criteria yet falls below the simplified rule; a false positive is the reverse. Sensitivity and specificity quantify these two error rates separately and are properties of the rule itself, largely stable across samples. Positive and negative predictive values translate the rule to a particular setting: because they depend on how common PTSD is, the same six-symptom definition yields different predictive values in a specialty clinic than in a community survey. Accuracy, the overall share of participants classified the same way as the full criteria, is the single number `score_by = "accuracy"` optimized for. Reporting all of these, alongside the underlying counts, is what lets a reader judge whether a simplified definition is adequate for their purpose.

`optimize_combinations()` returns these metrics for the reference diagnosis and each top combination in `res$summary`, including an `Accuracy` column. This is the diagnostic-performance table.

```{r summary}
res$summary
```

## Where next

- [Comparing diagnostic criteria](comparing-criteria.html) — evaluate several subset sizes and hierarchical rules against each other, add ICD-11 as a benchmark, and identify the symptoms that recur across solutions.
- [Validating abbreviated symptom definitions](validation.html) — test whether a definition holds up under cross-validation and in an independent cohort.
- [CAPS-5 workflow](caps5-workflow.html) — use the clinician-administered CAPS-5 as the reference instrument.
