---
title: "Joining diagnoses back to demographics with id_col"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Joining diagnoses back to demographics with id_col}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

## The problem

The `PTSDdiag` workflow normalises your dataframe to a canonical
`symptom_1` through `symptom_20` form before optimisation. Demographics
(`patient_id`, `age`, `sex`, …) are dropped at that step, so the
optimised diagnosis cannot be joined back to the original dataframe to
answer questions like *"do participants meeting the optimised criterion
differ in age from those who don't?"*

This vignette shows how to thread a participant identifier through the
workflow using the `id_col` argument so that per-row diagnoses can be
merged back to demographics.

## Set up a dataframe with demographics

```{r setup, message = FALSE}
library(PTSDdiag)
library(dplyr)

set.seed(123)
n <- 200
study_df <- data.frame(
  patient_id = sprintf("P%03d", seq_len(n)),
  age        = sample(20:70, n, replace = TRUE),
  sex        = sample(c("F", "M"), n, replace = TRUE),
  simulated_ptsd[seq_len(n), ],
  stringsAsFactors = FALSE
)
head(study_df[, 1:6])
```

## Carry the ID through `rename_ptsd_columns()`

Pass the column name(s) of any participant-level fields you want to
preserve. The simplest pattern is to use a single participant identifier
and join everything else back at the end:

```{r rename}
sym <- rename_ptsd_columns(study_df[, c("patient_id", names(simulated_ptsd))],
                           id_col = "patient_id")
names(sym)[1:5]
```

`sym` now contains `patient_id` plus `symptom_1` through `symptom_20`.
`age` and `sex` stay in `study_df` and will be joined back via
`patient_id` at the end. If you wanted them to ride through the workflow
as well, pass them too: `id_col = c("patient_id", "age", "sex")`.

## Optimise — the ID rides along

```{r optimize, message = FALSE}
res <- optimize_combinations(
  sym, n_symptoms = 6, n_required = 4, n_top = 2,
  show_progress = FALSE
)
head(res$diagnosis_comparison)
```

`res$diagnosis_comparison` carries `patient_id` as its first column.
Every diagnostic column (`PTSD_orig`, the optimised combinations) is
aligned to the same `patient_id`.

## Join diagnoses back to demographics

```{r join}
final <- study_df %>%
  select(patient_id, age, sex) %>%
  left_join(res$diagnosis_comparison, by = "patient_id")
head(final)
```

Now you can compare demographics by diagnostic status. The column
generated by the best optimised combination is named after its symptom
indices (e.g. `symptom_3_6_10_15_17_19`); pick the first non-`PTSD_orig`
diagnostic column for the demonstration:

```{r demographics}
combo_col <- setdiff(grep("^symptom_", names(final), value = TRUE),
                     paste0("symptom_", 1:20))[1]

final %>%
  group_by(.data[[combo_col]]) %>%
  summarise(n = n(), mean_age = mean(age), pct_female = mean(sex == "F"))
```

## The same trick works for validation

The ID rides through `holdout_validation()` and `cross_validation()` too
— useful because those functions reshuffle rows and positional `cbind`
would silently misalign your data.

```{r validate, message = FALSE}
ho <- holdout_validation(sym, train_ratio = 0.7, seed = 42,
                         n_symptoms = 6, n_required = 4, n_top = 2)
head(ho$without_clusters$test_results)
```

Each row of `test_results` carries the `patient_id` it came from, so a
`left_join` from demographics recovers the test fold exactly.

```{r validate-join}
study_df %>%
  select(patient_id, age, sex) %>%
  inner_join(ho$without_clusters$test_results, by = "patient_id") %>%
  head()
```

## Reserved column names

A small set of column names is reserved by the workflow and cannot be
used as `id_col`: `symptom_1` through `symptom_20`, `total`, `PTSD_orig`,
`PTSD_icd11`, `PTSD_caps5`, `PTSD_pcl5`, `.strata`, and any name matching
the combination-output pattern (e.g. `symptom_1_6_8_10_15_19`). Rename
your column before calling `rename_ptsd_columns()` if it collides.
