---
title: "Full Internal Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Full Internal Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# 1. Introduction

The `PTSDdiag` package provides tools for analysing and optimising PTSD diagnostic criteria using PCL-5 (PTSD Checklist for DSM-5) data. This vignette demonstrates a complete single-dataset workflow:

 * Import and prepare PCL-5 data
 * Calculate descriptive statistics and reliability metrics
 * Identify optimal symptom combinations (hierarchical and non-hierarchical)
 * Apply ICD-11 PTSD diagnostic criteria and compare diagnostic systems
 * Export and import optimised combinations
 * Validate models using holdout and cross-validation

For workflows involving independent validation datasets, see the [External Validation (PCL-5)](external_validation_pcl5.html) and [External Validation (CAPS-5)](external_validation_caps5.html) vignettes.

# 2. Setup

## 2.1. Installation

Install the released version from CRAN:
```{r Installation CRAN, eval=FALSE, include=TRUE}
install.packages("PTSDdiag")
```

Or the development version from GitHub:
```{r Installation GitHub, eval=FALSE, include=TRUE}
# install.packages("devtools")
devtools::install_github("TobiasRSpiller/PTSDdiag")
```

## 2.2. Loading packages

```{r Loading PTSDdiag, echo=TRUE, warning=FALSE}
library("PTSDdiag")
```

```{r Loading other, echo=TRUE, warning=FALSE}
library(psych)     # For reliability analysis
```

# 3. Data preparation

## 3.1. Loading the sample data

The package includes `simulated_ptsd`, a dataset of 5,000 simulated PCL-5 records. Each record rates 20 PTSD symptoms on a 5-point scale (0 = Not at all, 1 = A little bit, 2 = Moderately, 3 = Quite a bit, 4 = Extremely).

The PCL-5 organises symptoms into four DSM-5 clusters:

 * Symptoms 1--5: Criterion B (Intrusion)
 * Symptoms 6--7: Criterion C (Avoidance)
 * Symptoms 8--14: Criterion D (Negative alterations in cognitions and mood)
 * Symptoms 15--20: Criterion E (Alterations in arousal and reactivity)

Input data must contain numeric values 0--4 only, with no missing values, 20 columns (one per symptom), and row-wise observations.

```{r Loading sample data, echo=TRUE}
data("simulated_ptsd")
head(simulated_ptsd)
```

## 3.2. Standardising column names

All package functions expect columns named `symptom_1` through `symptom_20`. Use `rename_ptsd_columns()` to standardise:

```{r Renaming columns, echo=TRUE}
simulated_ptsd_renamed <- rename_ptsd_columns(simulated_ptsd)
names(simulated_ptsd_renamed)
```

## 3.3. Descriptive statistics

We calculate total scores, apply DSM-5 diagnostic criteria, and summarise:

```{r Basic Descriptive Statistics, echo=TRUE}
# Calculate total scores (range 0-80)
simulated_ptsd_total <- calculate_ptsd_total(simulated_ptsd_renamed)

# Apply DSM-5 diagnostic criteria
simulated_ptsd_total_diagnosed <- create_ptsd_diagnosis_nonbinarized(simulated_ptsd_total)

# Summary statistics
summarize_ptsd(simulated_ptsd_total_diagnosed)
```

## 3.4. Reliability analysis

Cronbach's alpha assesses the internal consistency of the PCL-5 items:

```{r Reliability Analysis, echo=TRUE}
cronbach <- psych::alpha(subset(simulated_ptsd_total_diagnosed, select = -total))
round(cronbach$total, 3)
```

# 4. Identifying optimal symptom combinations

Standard PTSD diagnosis requires evaluating 20 symptoms across four clusters. The goal is to identify reduced symptom combinations that preserve the polythetic structure of the diagnosis while maintaining diagnostic accuracy.

We examine two structural approaches: a **hierarchical** approach (requiring at least one symptom from each DSM-5 cluster) and a **non-hierarchical** approach (selecting symptoms purely by accuracy).

> **Note on computation time.** The examples below use 4-symptom combinations (C(20, 4) = 4,845 to evaluate), which run in seconds. A typical analysis uses 6-symptom combinations (C(20, 6) = 38,760), which takes 1--2 minutes with 5,000 observations.

## 4.1. Hierarchical analysis

The hierarchical approach requires at least one symptom from each DSM-5 cluster, maintaining the structural logic of the diagnostic criteria.

The `score_by` argument controls the optimisation target:

 * `"newly_nondiagnosed"` -- minimise false negatives (missed diagnoses)
 * `"false_cases"` -- minimise total misclassifications

```{r Optimal Hierarchical Symptom Combinations, echo=TRUE, message=FALSE}
best_combinations_hierarchical <- optimize_combinations_clusters(
  simulated_ptsd_renamed,
  n_symptoms = 4,
  n_required = 3,
  n_top      = 3,
  score_by   = "newly_nondiagnosed",
  clusters   = list(B = 1:5, C = 6:7, D = 8:14, E = 15:20),
  show_progress = FALSE
)
```

The function returns three elements. **1. The best symptom combinations:**

```{r Best Symptoms hierarchical, echo=TRUE}
best_combinations_hierarchical$best_symptoms
```

**2. A row-level comparison of DSM-5-TR vs. each combination** (first 5 rows):

```{r Comparison of diagnosis hierarchical, echo=TRUE}
knitr::kable(
  head(best_combinations_hierarchical$diagnosis_comparison, 5),
  caption = "Diagnosis comparison: DSM-5-TR vs. top hierarchical combinations (first 5 rows)"
)
```

**3. Diagnostic accuracy metrics:**

```{r Summary table hierachical, echo=TRUE}
knitr::kable(
  best_combinations_hierarchical$summary,
  digits  = 4,
  caption = "Hierarchical combinations: diagnostic accuracy metrics"
)
```

## 4.2. Non-hierarchical analysis

The non-hierarchical approach selects symptoms purely by accuracy, without requiring cluster representation.

```{r Optimal Non-hierarchical Symptom Combinations, echo=TRUE, message=FALSE}
best_combinations_nonhierarchical <- optimize_combinations(
  simulated_ptsd_renamed,
  n_symptoms = 4,
  n_required = 3,
  n_top      = 3,
  score_by   = "newly_nondiagnosed",
  show_progress = FALSE
)
```

**Best symptom combinations:**

```{r Best Symptoms nonhierarchical, echo=TRUE}
best_combinations_nonhierarchical$best_symptoms
```

**Diagnosis comparison** (first 5 rows):

```{r Comparison of diagnosis nonhierarchical, echo=TRUE}
knitr::kable(
  head(best_combinations_nonhierarchical$diagnosis_comparison, 5),
  caption = "Diagnosis comparison: DSM-5-TR vs. top non-hierarchical combinations (first 5 rows)"
)
```

**Diagnostic accuracy metrics:**

```{r Summary table nonhierachical, echo=TRUE}
knitr::kable(
  best_combinations_nonhierarchical$summary,
  digits  = 4,
  caption = "Non-hierarchical combinations: diagnostic accuracy metrics"
)
```

# 5. Comparing diagnostic systems

Optimised combinations can be benchmarked against established reference standards: the full DSM-5-TR criteria and the ICD-11 PTSD definition.

## 5.1. ICD-11 PTSD criteria

The ICD-11 defines PTSD using a narrower symptom set than DSM-5-TR. Using PCL-5 items, the ICD-11 requires **all three** of the following clusters to be met (symptom present = score >= 2):

| ICD-11 Cluster | Requirement | PCL-5 Items |
|---|---|---|
| Re-experiencing (in the present) | >= 1 of items 1, 2, 3 | Intrusive memories, nightmares, flashbacks |
| Avoidance | >= 1 of items 6, 7 | Avoidance of thoughts, avoidance of reminders |
| Sense of current threat | >= 1 of items 16, 17 | Hypervigilance, exaggerated startle |

`create_icd11_diagnosis()` applies these criteria:

```{r ICD-11 diagnosis, echo=TRUE}
icd11_result <- create_icd11_diagnosis(simulated_ptsd_renamed)
head(icd11_result)
```

```{r ICD-11 prevalence, echo=TRUE}
cat(sprintf("DSM-5-TR prevalence: %.1f%%\n", mean(icd11_result$PTSD_orig)  * 100))
cat(sprintf("ICD-11 prevalence:   %.1f%%\n", mean(icd11_result$PTSD_icd11) * 100))
```

Diagnostic accuracy of ICD-11 relative to DSM-5-TR:

```{r ICD-11 metrics, echo=TRUE}
knitr::kable(
  create_readable_summary(summarize_ptsd_changes(icd11_result)),
  digits  = 4,
  caption = "ICD-11 vs. DSM-5-TR: diagnostic accuracy metrics"
)
```

## 5.2. Unified comparison table

`compare_diagnostic_systems()` combines any number of diagnostic criteria into a single table. Here we compare DSM-5-TR, ICD-11, and both sets of optimised combinations:

```{r Compare with optimization output, echo=TRUE, message=FALSE}
tbl_full <- compare_diagnostic_systems(
  simulated_ptsd_renamed,
  best_combinations_nonhierarchical$diagnosis_comparison,
  best_combinations_hierarchical$diagnosis_comparison,
  icd11 = TRUE
)

knitr::kable(
  tbl_full,
  digits  = 4,
  caption = "Comparison: DSM-5-TR, ICD-11, non-hierarchical, and hierarchical combinations"
)
```

Use the `labels` argument to rename the optimised combination rows (by default, column names such as `symptom_1_6_8_15` are used). Set `icd11 = FALSE` to omit the ICD-11 row.

# 6. Exporting and importing combinations

`write_combinations()` and `read_combinations()` export and import optimised combinations as human-readable JSON files, enabling sharing across research groups.

## 6.1. Saving derived combinations

```{r Export combinations intro, echo=TRUE, message=FALSE}
tmp_hier    <- tempfile(fileext = ".json")
tmp_nonhier <- tempfile(fileext = ".json")

write_combinations(
  best_combinations_hierarchical$best_symptoms,
  file        = tmp_hier,
  n_required  = 3,
  clusters    = list(B = 1:5, C = 6:7, D = 8:14, E = 15:20),
  score_by    = "newly_nondiagnosed",
  description = "Top 3 hierarchical combinations from simulated data"
)

write_combinations(
  best_combinations_nonhierarchical$best_symptoms,
  file       = tmp_nonhier,
  n_required = 3,
  score_by   = "newly_nondiagnosed",
  description = "Top 3 non-hierarchical combinations from simulated data"
)

cat("Files written successfully.\n")
```

## 6.2. Loading saved combinations

A collaborator can load the combinations and apply them to new data:

```{r Import combinations intro, echo=TRUE, message=FALSE}
spec <- read_combinations(tmp_hier)

cat("Combinations loaded: ", length(spec$combinations), "\n")
cat("n_required:          ", spec$n_required, "\n")
cat("Clusters:            ", paste(names(spec$clusters), collapse = ", "), "\n")
cat("Description:         ", spec$description, "\n")
```

```{r Apply loaded combinations, echo=TRUE, message=FALSE}
comparison_loaded <- apply_symptom_combinations(
  simulated_ptsd_renamed,
  combinations = spec$combinations,
  n_required   = spec$n_required,
  clusters     = spec$clusters
)

knitr::kable(
  create_readable_summary(summarize_ptsd_changes(comparison_loaded)),
  digits  = 4,
  caption = "Performance of re-loaded hierarchical combinations"
)
```

For a complete derivation-validation workflow across independent datasets, see the [External Validation (PCL-5)](external_validation_pcl5.html) vignette.

# 7. Model validation

The package provides two complementary approaches for estimating how well optimised combinations generalise to new data.

> **Note on computation time.** The examples below use `n_symptoms = 4` to keep build time short. Real-world analyses typically use `n_symptoms = 6`.

## 7.1. Holdout validation

Holdout validation splits the data into training (70%) and test (30%) sets. Optimal combinations are derived on the training set and evaluated on the held-out test set.

```{r Holdout Validation, echo=TRUE, message=FALSE}
validation_results <- holdout_validation(
  simulated_ptsd_renamed,
  train_ratio = 0.7,
  score_by    = "newly_nondiagnosed",
  seed        = 123,
  n_symptoms  = 4,
  n_required  = 3,
  n_top       = 3
)
```

**Best combinations identified on training data:**

```{r Holdout best combinations, echo=TRUE}
cat("Non-hierarchical:\n")
print(validation_results$without_clusters$best_combinations)
cat("\nHierarchical:\n")
print(validation_results$with_clusters$best_combinations)
```

**Test-set performance:**

```{r Holdout Results Non-Hierarchical, echo=TRUE}
knitr::kable(
  validation_results$without_clusters$summary,
  digits  = 4,
  caption = "Holdout validation: non-hierarchical combinations (test set)"
)
```

```{r Holdout Results Hierarchical, echo=TRUE}
knitr::kable(
  validation_results$with_clusters$summary,
  digits  = 4,
  caption = "Holdout validation: hierarchical combinations (test set)"
)
```

## 7.2. Cross-validation

k-fold cross-validation provides a more robust estimate of generalisation by repeating the train/test split `k` times.

```{r Cross Validation, echo=TRUE, message=FALSE}
cv_results <- cross_validation(
  simulated_ptsd_renamed,
  k          = 3,
  score_by   = "newly_nondiagnosed",
  seed       = 123,
  n_symptoms = 4,
  n_required = 3,
  n_top      = 3
)
```

**Results by fold -- non-hierarchical:**

```{r CV Results by Fold for non-hierarchical model, echo=TRUE}
knitr::kable(
  cv_results$without_clusters$summary_by_fold,
  digits  = 4,
  caption = "Cross-validation: non-hierarchical results by fold"
)
```

**Results by fold -- hierarchical:**

```{r CV Results by Fold for hierarchical model, echo=TRUE}
knitr::kable(
  cv_results$with_clusters$summary_by_fold,
  digits  = 4,
  caption = "Cross-validation: hierarchical results by fold"
)
```

**Stable combinations across folds** -- if a combination appeared in multiple folds, its average performance is reported:

```{r CV Stable Combinations for non-hierarchical model, echo=TRUE}
if (!is.null(cv_results$without_clusters$combinations_summary)) {
  knitr::kable(
    cv_results$without_clusters$combinations_summary,
    digits  = 4,
    caption = "Stable non-hierarchical combinations across folds"
  )
} else {
  cat("No combination appeared in more than one fold (non-hierarchical).\n")
}
```

```{r CV Stable Combinations for hierarchical model, echo=TRUE}
if (!is.null(cv_results$with_clusters$combinations_summary)) {
  knitr::kable(
    cv_results$with_clusters$combinations_summary,
    digits  = 4,
    caption = "Stable hierarchical combinations across folds"
  )
} else {
  cat("No combination appeared in more than one fold (hierarchical).\n")
}
```

## 7.3. Comparing validation approaches

| | Holdout | Cross-Validation |
|---|---|---|
| Train/test splits | 1 | k |
| Speed | Fast | Slower |
| Best for | Quick assessment | Publication-quality results |
| Stability estimate | Single estimate | Average over k folds |

```{r Compare Validation Methods, echo=TRUE}
holdout_sens <- validation_results$without_clusters$summary$Sensitivity[2]
cv_fold_sens <- cv_results$without_clusters$summary_by_fold$sensitivity

cat(sprintf("Holdout sensitivity (top non-hierarchical combination): %.4f\n", holdout_sens))
cat(sprintf("CV mean sensitivity across folds:                      %.4f\n", mean(cv_fold_sens, na.rm = TRUE)))
cat(sprintf("CV sensitivity range:                                  %.4f -- %.4f\n",
            min(cv_fold_sens, na.rm = TRUE), max(cv_fold_sens, na.rm = TRUE)))
```

## 7.4. Best practices

1. Set a seed for reproducibility: `seed = 123`
2. Use holdout for initial assessment; cross-validation for publication-quality results
3. Choose an appropriate optimisation target:
   - `score_by = "newly_nondiagnosed"` to minimise missed diagnoses
   - `score_by = "false_cases"` to minimise total misclassifications
4. Examine multiple metrics: sensitivity for screening, specificity for differential diagnosis, PPV/NPV for clinical decision-making

# 8. Conclusion

For workflows that test whether optimised criteria generalise to independent samples, see:

 * [External Validation (PCL-5)](external_validation_pcl5.html) -- derivation on one PCL-5 dataset, validation on another
 * [External Validation (CAPS-5)](external_validation_caps5.html) -- validation against the CAPS-5 gold standard
