---
title: "An introduction to PTSDdiag"
output: rmarkdown::html_document
vignette: >
  %\VignetteIndexEntry{An introduction to PTSDdiag}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# 1. Introduction
The `PTSDdiag` package provides tools for analyzing and optimizing PTSD diagnostic criteria using PCL-5 (PTSD Checklist for DSM-5) data. Post-Traumatic Stress Disorder (PTSD) diagnosis traditionally requires a complex evaluation of multiple symptom criteria across different clusters. This package aims to simplify this process while maintaining diagnostic accuracy.

This vignette demonstrates how to use the package to:

 * Import and prepare PCL-5 data
 * Calculate basic descriptive statistics and reliability metrics
 * Find optimal symptom combinations for PTSD diagnosis
 * Apply ICD-11 PTSD diagnostic criteria
 * Compare diagnostic systems (DSM-5-TR, ICD-11, and optimized combinations) in a unified table
 * Export and import optimized combinations
 * Validate models using holdout and cross-validation

# 2. Installation
## 2.1. Installing the Package

You can install the released version from CRAN:
```{r Installation CRAN, eval=FALSE, include=TRUE}
install.packages("PTSDdiag")
```

Or install the development version from GitHub:
```{r Installation GitHub, eval=FALSE, include=TRUE}
# install.packages("devtools")
devtools::install_github("TobiasRSpiller/PTSDdiag")
```

## 2.2. Loading Required Packages

Once the `PTSDdiag` is installed, it can be loaded the usual way.
```{r Loading PTSDdiag, echo=TRUE, warning=FALSE}
library("PTSDdiag")
```

Load additional packages:
```{r Loading other, echo=TRUE, warning=FALSE}
library(psych)     # For reliability analysis
```

# 3. Basic Usage
## 3.1. Loading Sample Data
This package includes a simulated dataset that mirrors hypothetical PCL-5-assessments. It contains 5,000 simulated patient responses, each rating 20 PTSD symptoms according to DSM-5 criteria.

**Rating Scale**

Each symptom is rated on a 5-point scale:

 * 0 = Not at all
 * 1 = A little bit
 * 2 = Moderately
 * 3 = Quite a bit
 * 4 = Extremely

**Symptom Clusters**

The PCL-5 organizes symptoms into four distinct clusters according to DSM-5 criteria:

 * Symptoms 1-5: Criterion B (Intrusion)
 * Symptoms 6-7: Criterion C (Avoidance)
 * Symptoms 8-14: Criterion D (Negative alterations in cognitions and mood)
 * Symptoms 15-20: Criterion E (Alterations in arousal and reactivity)

**Data Format Requirements**

Input data must be:

 * Numeric values 0-4 only
 * No missing values
 * 20 columns (one per symptom)
 * Row-wise observations

Let's load the included sample data:
```{r Loading sample data, echo=TRUE}
data("simulated_ptsd")
head(simulated_ptsd)
```

## 3.2. Data Preparation
The first step is to standardize column names for consistent analysis. Before standardization, columns might have various names (here `S1`–`S20`). After standardization they follow the `symptom_1`–`symptom_20` pattern expected by all package functions:

```{r Renaming columns, echo=TRUE}
simulated_ptsd_renamed <- rename_ptsd_columns(simulated_ptsd)
names(simulated_ptsd_renamed)
```

## 3.3. Basic Descriptive Statistics
We process the data through three steps: calculate total scores, apply DSM-5 diagnostic criteria, and summarize.

```{r Basic Descriptive Statistics, echo=TRUE}
# Step 1: Calculate total scores (range 0-80)
simulated_ptsd_total <- calculate_ptsd_total(simulated_ptsd_renamed)

# Step 2: Apply DSM-5 diagnostic criteria
simulated_ptsd_total_diagnosed <- create_ptsd_diagnosis_nonbinarized(simulated_ptsd_total)

# Step 3: Summary statistics
summarize_ptsd(simulated_ptsd_total_diagnosed)
```

The summary provides mean total score, standard deviation, and number of DSM-5-diagnosed cases.

## 3.4. Reliability Analysis
Cronbach's alpha assesses the internal consistency of the PCL-5 items:

```{r Reliability Analysis, echo=TRUE}
cronbach <- psych::alpha(subset(simulated_ptsd_total_diagnosed, select = -total))
round(cronbach$total, 3)
```

# 4. Finding Optimal Symptom Combinations

Current PTSD diagnosis requires evaluating 20 symptoms across four clusters with complex rules. Our goal is to identify simplified diagnostic criteria that:

 * Reduce the number of symptoms to evaluate
 * Preserve the polythetic nature of the diagnosis (a minimum number of the selected symptoms must be present)
 * Maintain diagnostic accuracy compared to the original DSM-5 criteria

We identify optimal symptom combinations under two structural approaches: a **hierarchical** approach (requiring at least one symptom from each DSM-5 cluster) and a **non-hierarchical** approach (ignoring cluster membership).

> **Note on computation time.** The examples below use 4-symptom combinations (C(20, 4) = 4,845 to evaluate), which run in seconds. A typical real-world analysis uses 6-symptom combinations (C(20, 6) = 38,760), which takes 1–2 minutes per call with 5,000 observations. The workflow and output structure are identical regardless of `n_symptoms`.

## 4.1. Hierarchical Analysis

In the hierarchical approach at least one symptom from each DSM-5 cluster must be included in the combination. This maintains the structural logic of the diagnostic criteria.

The `score_by` argument controls what "optimal" means:

 * `"newly_nondiagnosed"` — minimize false negatives (missed diagnoses)
 * `"false_cases"` — minimize total misclassifications

```{r Optimal Hierarchical Symptom Combinations, echo=TRUE, message=FALSE}
best_combinations_hierarchical <- optimize_combinations_clusters(
  simulated_ptsd_renamed,
  n_symptoms = 4,
  n_required = 3,
  n_top      = 3,
  score_by   = "newly_nondiagnosed",
  clusters   = list(B = 1:5, C = 6:7, D = 8:14, E = 15:20),
  show_progress = FALSE
)
```

The function returns three elements. **1. The best symptom combinations identified:**

```{r Best Symptoms hierarchical, echo=TRUE}
best_combinations_hierarchical$best_symptoms
```

**2. A row-level comparison of DSM-5-TR vs. each new combination** (first 5 rows shown):

```{r Comparison of diagnosis hierarchical, echo=TRUE}
knitr::kable(
  head(best_combinations_hierarchical$diagnosis_comparison, 5),
  caption = "Diagnosis comparison: DSM-5-TR vs. top hierarchical combinations (first 5 rows)"
)
```

**3. Diagnostic accuracy metrics for each combination:**

```{r Summary table hierachical, echo=TRUE}
knitr::kable(
  best_combinations_hierarchical$summary,
  digits  = 4,
  caption = "Hierarchical combinations: diagnostic accuracy metrics"
)
```

The summary includes total diagnosed/non-diagnosed counts, true/false positives and negatives, and sensitivity, specificity, PPV, and NPV.

## 4.2. Non-Hierarchical Analysis

The non-hierarchical approach selects symptoms purely by accuracy, without requiring cluster representation.

```{r Optimal Non-hierarchical Symptom Combinations, echo=TRUE, message=FALSE}
best_combinations_nonhierarchical <- optimize_combinations(
  simulated_ptsd_renamed,
  n_symptoms = 4,
  n_required = 3,
  n_top      = 3,
  score_by   = "newly_nondiagnosed",
  show_progress = FALSE
)
```

**Best symptom combinations:**

```{r Best Symptoms nonhierarchical, echo=TRUE}
best_combinations_nonhierarchical$best_symptoms
```

**Diagnosis comparison** (first 5 rows):

```{r Comparison of diagnosis nonhierarchical, echo=TRUE}
knitr::kable(
  head(best_combinations_nonhierarchical$diagnosis_comparison, 5),
  caption = "Diagnosis comparison: DSM-5-TR vs. top non-hierarchical combinations (first 5 rows)"
)
```

**Diagnostic accuracy metrics:**

```{r Summary table nonhierachical, echo=TRUE}
knitr::kable(
  best_combinations_nonhierarchical$summary,
  digits  = 4,
  caption = "Non-hierarchical combinations: diagnostic accuracy metrics"
)
```

# 5. Comparing Diagnostic Systems

After identifying optimal symptom combinations, a natural next step is to benchmark them against established reference standards: the full DSM-5-TR criteria and the ICD-11 PTSD definition. The `PTSDdiag` package provides two functions for this purpose.

## 5.1. ICD-11 PTSD Criteria

The ICD-11 defines PTSD using a narrower and more parsimonious symptom set than DSM-5-TR. Using PCL-5 items, the ICD-11 requires **all three** of the following clusters to be met (symptom present = score ≥ 2):

| ICD-11 Cluster | Requirement | PCL-5 Items |
|---|---|---|
| Re-experiencing (in the present) | ≥ 1 of items 1, 2, 3 | Intrusive memories, nightmares, flashbacks |
| Avoidance | ≥ 1 of items 6, 7 | Avoidance of thoughts, avoidance of reminders |
| Sense of current threat | ≥ 1 of items 16, 17 | Hypervigilance, exaggerated startle |

`create_icd11_diagnosis()` applies these criteria and returns a two-column data frame (`PTSD_orig`, `PTSD_icd11`) that is directly compatible with the existing metrics pipeline.

```{r ICD-11 diagnosis, echo=TRUE}
icd11_result <- create_icd11_diagnosis(simulated_ptsd_renamed)
head(icd11_result)
```

```{r ICD-11 prevalence, echo=TRUE}
cat(sprintf("DSM-5-TR prevalence: %.1f%%\n", mean(icd11_result$PTSD_orig)  * 100))
cat(sprintf("ICD-11 prevalence:   %.1f%%\n", mean(icd11_result$PTSD_icd11) * 100))
```

The output feeds directly into the standard metrics pipeline:

```{r ICD-11 metrics, echo=TRUE}
knitr::kable(
  create_readable_summary(summarize_ptsd_changes(icd11_result)),
  digits  = 4,
  caption = "ICD-11 vs. DSM-5-TR: diagnostic accuracy metrics"
)
```

## 5.2. Unified Comparison Table

`compare_diagnostic_systems()` combines any number of diagnostic criteria — DSM-5-TR (always the reference), ICD-11 (computed internally when `icd11 = TRUE`), and any pre-specified or optimized combinations — into a single presentation-ready table.

**ICD-11 vs. DSM-5-TR only:**

```{r Compare systems ICD-11 only, echo=TRUE}
knitr::kable(
  compare_diagnostic_systems(simulated_ptsd_renamed, icd11 = TRUE),
  digits  = 4,
  caption = "Comparison table: DSM-5-TR and ICD-11"
)
```

**Adding optimized combinations:**

Pass comparison data frames from `apply_symptom_combinations()` alongside the raw data. You can equally pass the `$diagnosis_comparison` slot produced by `optimize_combinations()` or `optimize_combinations_clusters()` — it shares the same structure. Here we reuse the derivation results from Section 4:

```{r Compare with optimization output, echo=TRUE, message=FALSE}
tbl_full <- compare_diagnostic_systems(
  simulated_ptsd_renamed,
  best_combinations_nonhierarchical$diagnosis_comparison,
  best_combinations_hierarchical$diagnosis_comparison,
  icd11 = TRUE
)

knitr::kable(
  tbl_full,
  digits  = 4,
  caption = "Comparison table: DSM-5-TR, ICD-11, non-hierarchical and hierarchical combinations"
)
```

**Key points:**

 * `"DSM-5-TR"` and `"ICD-11"` labels are applied automatically — no need to specify them in `labels`.
 * Use the `labels` argument to rename the optimized combination rows (by default, column names such as `symptom_1_6_8_15` are used).
 * Set `icd11 = FALSE` to omit the ICD-11 row if not needed.
 * The function validates that all `PTSD_orig` columns match the reference computed from `data`, ensuring consistency.

# 6. Exporting and Importing Combinations

Once you have identified optimal symptom combinations, you can share them with collaborators or save them for later use. `write_combinations()` and `read_combinations()` export and import combinations as human-readable JSON files.

## 6.1. Saving Derived Combinations

```{r Export combinations intro, echo=TRUE, message=FALSE}
# Save to a temporary file for this demonstration
# In practice, specify a permanent path: file = "hierarchical_combos.json"
tmp_hier    <- tempfile(fileext = ".json")
tmp_nonhier <- tempfile(fileext = ".json")

write_combinations(
  best_combinations_hierarchical$best_symptoms,
  file        = tmp_hier,
  n_required  = 3,
  clusters    = list(B = 1:5, C = 6:7, D = 8:14, E = 15:20),
  score_by    = "newly_nondiagnosed",
  description = "Top 3 hierarchical combinations from simulated data"
)

write_combinations(
  best_combinations_nonhierarchical$best_symptoms,
  file       = tmp_nonhier,
  n_required = 3,
  score_by   = "newly_nondiagnosed",
  description = "Top 3 non-hierarchical combinations from simulated data"
)

cat("Files written successfully.\n")
```

## 6.2. Loading Saved Combinations

A collaborator (or you in a future session) can load the combinations and apply them to new data:

```{r Import combinations intro, echo=TRUE, message=FALSE}
# Read the exported file
spec <- read_combinations(tmp_hier)

# Inspect the stored metadata
cat("Combinations loaded: ", length(spec$combinations), "\n")
cat("n_required:          ", spec$n_required, "\n")
cat("Clusters:            ", paste(names(spec$clusters), collapse = ", "), "\n")
cat("Description:         ", spec$description, "\n")
```

```{r Apply loaded combinations, echo=TRUE, message=FALSE}
# Apply to any dataset with the same column structure
comparison_loaded <- apply_symptom_combinations(
  simulated_ptsd_renamed,
  combinations = spec$combinations,
  n_required   = spec$n_required,
  clusters     = spec$clusters
)

knitr::kable(
  create_readable_summary(summarize_ptsd_changes(comparison_loaded)),
  digits  = 4,
  caption = "Performance of re-loaded hierarchical combinations"
)
```

This is particularly useful for external validation workflows where one research group derives the criteria and another validates them on an independent dataset. See the *External Validation* vignette for a full worked example.

# 7. Model Validation
After identifying optimal symptom combinations, it is crucial to validate their performance on independent data. The `PTSDdiag` package provides two validation approaches.

> **Note on computation time.** The examples below use `n_symptoms = 4` to keep build time short. Real-world analyses typically use `n_symptoms = 6`.

## 7.1. Holdout Validation
Holdout validation splits the data into training (70%) and test (30%) sets. Optimal combinations are derived on the training set and evaluated on the held-out test set.

```{r Holdout Validation, echo=TRUE, message=FALSE}
validation_results <- holdout_validation(
  simulated_ptsd_renamed,
  train_ratio = 0.7,
  score_by    = "newly_nondiagnosed",
  seed        = 123,
  n_symptoms  = 4,
  n_required  = 3,
  n_top       = 3
)
```

**Best combinations identified on training data:**

```{r Holdout best combinations, echo=TRUE}
cat("Non-hierarchical:\n")
print(validation_results$without_clusters$best_combinations)
cat("\nHierarchical:\n")
print(validation_results$with_clusters$best_combinations)
```

**Test-set performance:**

```{r Holdout Results Non-Hierarchical, echo=TRUE}
knitr::kable(
  validation_results$without_clusters$summary,
  digits  = 4,
  caption = "Holdout validation: non-hierarchical combinations (test set)"
)
```

```{r Holdout Results Hierarchical, echo=TRUE}
knitr::kable(
  validation_results$with_clusters$summary,
  digits  = 4,
  caption = "Holdout validation: hierarchical combinations (test set)"
)
```

## 7.2. Cross-Validation
k-fold cross-validation provides a more robust estimate of generalization by repeating the train/test split `k` times.

```{r Cross Validation, echo=TRUE, message=FALSE}
cv_results <- cross_validation(
  simulated_ptsd_renamed,
  k          = 3,
  score_by   = "newly_nondiagnosed",
  seed       = 123,
  n_symptoms = 4,
  n_required = 3,
  n_top      = 3
)
```

**Results by fold — non-hierarchical:**

```{r CV Results by Fold for non-hierarchical model, echo=TRUE}
knitr::kable(
  cv_results$without_clusters$summary_by_fold,
  digits  = 4,
  caption = "Cross-validation: non-hierarchical results by fold"
)
```

**Results by fold — hierarchical:**

```{r CV Results by Fold for hierarchical model, echo=TRUE}
knitr::kable(
  cv_results$with_clusters$summary_by_fold,
  digits  = 4,
  caption = "Cross-validation: hierarchical results by fold"
)
```

**Stable combinations across folds** — if a combination appeared in multiple folds, its average performance is reported:

```{r CV Stable Combinations for non-hierarchical model, echo=TRUE}
if (!is.null(cv_results$without_clusters$combinations_summary)) {
  knitr::kable(
    cv_results$without_clusters$combinations_summary,
    digits  = 4,
    caption = "Stable non-hierarchical combinations across folds"
  )
} else {
  cat("No combination appeared in more than one fold (non-hierarchical).\n")
}
```

```{r CV Stable Combinations for hierarchical model, echo=TRUE}
if (!is.null(cv_results$with_clusters$combinations_summary)) {
  knitr::kable(
    cv_results$with_clusters$combinations_summary,
    digits  = 4,
    caption = "Stable hierarchical combinations across folds"
  )
} else {
  cat("No combination appeared in more than one fold (hierarchical).\n")
}
```

**Interpreting cross-validation results:**

1. **Consistency** — whether the same combinations are selected across folds
2. **Stability** — how diagnostic accuracy varies across folds
3. **Generalization** — average metrics provide robust real-world estimates

Key metrics: `Splits_Appeared` (how many folds selected a combination), sensitivity, specificity, PPV, and NPV.

## 7.3. Comparing Validation Approaches

Both methods serve complementary purposes:

| | Holdout | Cross-Validation |
|---|---|---|
| Train/test splits | 1 | k |
| Speed | Fast | Slower |
| Best for | Quick assessment | Publication-quality results |
| Stability estimate | Single estimate | Average over k folds |

```{r Compare Validation Methods, echo=TRUE}
# Extract sensitivity for the top combination from each approach
holdout_sens <- validation_results$without_clusters$summary$Sensitivity[2]
cv_fold_sens <- cv_results$without_clusters$summary_by_fold$sensitivity

cat(sprintf("Holdout sensitivity (top non-hierarchical combo): %.4f\n", holdout_sens))
cat(sprintf("CV mean sensitivity across folds:                 %.4f\n", mean(cv_fold_sens, na.rm = TRUE)))
cat(sprintf("CV sensitivity range:                             %.4f – %.4f\n",
            min(cv_fold_sens, na.rm = TRUE), max(cv_fold_sens, na.rm = TRUE)))
```

## 7.4. Validation Best Practices

1. Set a seed for reproducibility: `seed = 123`
2. Choose the appropriate method:
   - Use holdout for a quick initial assessment
   - Use cross-validation for publication-quality results
3. Consider your optimization goal:
   - `score_by = "newly_nondiagnosed"` to minimize missed diagnoses
   - `score_by = "false_cases"` to minimize total misclassifications
4. Examine multiple metrics — sensitivity for screening, specificity for differential diagnosis, PPV/NPV for clinical decision-making

# 8. Conclusion
With the `PTSDdiag` package, PCL-5 data can be processed and analyzed efficiently. It allows researchers to identify reduced optimal symptom combinations for PTSD diagnosis, apply ICD-11 criteria, and compare multiple diagnostic systems — DSM-5-TR, ICD-11, and optimized combinations — in a single unified table. Holdout and cross-validation functions provide robust estimates of how well the derived criteria generalize to new samples.
