---
title: "Multi-scenario optimization analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multi-scenario optimization analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 7,
  fig.height = 4
)
```

## Goal

This vignette replicates the multi-scenario derivation workflow used in the
PTSDdiag preprint in roughly twenty lines of code. Three optimization
scenarios are compared on the same sample — 4-of-6 hierarchical, 4-of-6
non-hierarchical, and 3-of-6 non-hierarchical — and the results are summarised
as a manuscript-ready table plus a symptom-selection heatmap that helps
identify "core" symptoms recurring across the data-driven combinations.

```{r setup, message = FALSE}
library(PTSDdiag)
```

## Estimate all scenarios with one call

`compare_optimizations()` runs every scenario for you. By default it uses the
three preprint scenarios. Set `include_icd11 = TRUE` to add ICD-11 as a
benchmark fixed criterion in the same comparison.

```{r estimate, message = FALSE}
ptsd_data <- rename_ptsd_columns(simulated_ptsd[1:500, ])

comp <- compare_optimizations(
  ptsd_data,
  n_top         = 10,
  include_icd11 = TRUE,
  show_progress = FALSE
)
print(comp)
```

## Manuscript Table 2

`summarize_top_combinations()` extracts the per-combination performance
metrics for each scenario in one tidy data frame. Set `as_percent = TRUE` to
present Sensitivity / Specificity / PPV / NPV as percentages, and `top_n` to
limit the per-scenario rows shown in the manuscript table (the preprint uses
the top 5).

```{r table-2}
tbl <- summarize_top_combinations(comp, top_n = 5, as_percent = TRUE)
head(tbl, 12)
```

Pipe the result into your favourite presentation table package, e.g.

```r
flextable::flextable(tbl) |> flextable::theme_vanilla()
```

## Figure 1 — symptom-selection heatmap

`plot_symptom_frequency()` returns a `ggplot` object. The OVERALL row pools
across the optimization scenarios so you can see which symptoms recur most.
Fixed criteria such as ICD-11 appear with 100% inclusion on their fixed
symptoms and 0% elsewhere.

```{r figure-1, fig.alt = "Heatmap of PCL-5 symptom selection frequency across optimization scenarios"}
plot_symptom_frequency(comp, type = "relative")
```

You can extend the plot with the usual `+` operator:

```r
plot_symptom_frequency(comp) +
  ggplot2::ggtitle("My title") +
  ggplot2::theme(text = ggplot2::element_text(size = 12))
```

## Supplementary Table S4 — raw counts per symptom

```{r table-s4}
freq <- symptom_frequency(comp)
head(freq, 12)
```

Pivot to a wide table per scenario in one line:

```{r table-s4-wide}
if (requireNamespace("tidyr", quietly = TRUE)) {
  tidyr::pivot_wider(freq[freq$Approach != "OVERALL", ],
                     names_from  = Approach,
                     values_from = Count) |>
    head()
}
```

## Customising scenarios

`scenarios` is a named list of configurations. You can vary
`n_symptoms`, `n_required`, and `hierarchical` freely — including
`n_symptoms = 5` or `7` — and mix in fixed criteria via `type = "fixed"`.

```{r custom-scenarios, eval = FALSE}
my_scenarios <- list(
  "5/7 Hierarchical"      = list(n_symptoms = 7, n_required = 5, hierarchical = TRUE),
  "4/6 Hierarchical"      = list(n_symptoms = 6, n_required = 4, hierarchical = TRUE),
  "4/6 Non-hierarchical"  = list(n_symptoms = 6, n_required = 4, hierarchical = FALSE),
  "ICD-11"                = list(type = "fixed", criterion = "icd11")
)
compare_optimizations(ptsd_data, scenarios = my_scenarios, n_top = 10)
```

## Joining diagnoses back to demographics

`compare_optimizations()` honours the `id_col` carry-through: if you supply
participant identifiers via `rename_ptsd_columns(..., id_col = "patient_id")`,
they are prepended to every scenario's per-row `diagnosis_comparison`. You
can then `dplyr::left_join()` to your original dataframe and analyse
demographics by diagnostic status. See the
[Joining diagnoses to demographics](id_column_workflow.html) vignette for the
full pattern.
