---
title: "Quick Start"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quick Start}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette gets you from installation to a first optimised PTSD diagnostic combination in about 5 minutes. For the full workflow — reliability analysis, ICD-11 comparison, model validation — see the [Full Internal Analysis](internal_analysis.html) vignette.

## Install & load

```{r install, eval = FALSE}
# From CRAN
install.packages("PTSDdiag")

# Or the development version from GitHub
# install.packages("devtools")
devtools::install_github("TobiasRSpiller/PTSDdiag")
```

```{r load, message = FALSE}
library(PTSDdiag)
```

## Load sample data

The package ships with `simulated_ptsd`: 5,000 simulated PCL-5 records (20 symptoms, 0–4 scale), structured to reflect a trauma-exposed clinical sample.

```{r data}
data("simulated_ptsd")
dim(simulated_ptsd)   # 5000 rows, 20 columns
head(simulated_ptsd[, 1:5])
```

## Prepare & score

Standardise column names to the `symptom_1`–`symptom_20` convention, then calculate total scores and apply DSM-5-TR criteria:

```{r prepare}
df <- rename_ptsd_columns(simulated_ptsd)
df <- calculate_ptsd_total(df)
df <- create_ptsd_diagnosis_nonbinarized(df)

summarize_ptsd(df)
```

## Find an optimal symptom combination

`optimize_combinations()` searches for the best *k*-symptom combinations (here: 4 out of 20) that preserve DSM-5-TR diagnostic accuracy. Setting `score_by = "newly_nondiagnosed"` minimises missed diagnoses (false negatives).

> **Note on computation time.** This example uses `n_symptoms = 4` (4,845 combinations) for speed. A typical research analysis uses `n_symptoms = 6` (38,760 combinations), which takes 1–2 minutes.

```{r optimize, message = FALSE}
result <- optimize_combinations(
  rename_ptsd_columns(simulated_ptsd),
  n_symptoms    = 4,
  n_required    = 3,
  n_top         = 3,
  score_by      = "newly_nondiagnosed",
  show_progress = FALSE
)
```

**Top combinations found:**

```{r best-symptoms}
result$best_symptoms
```

**Diagnostic accuracy metrics:**

```{r summary}
knitr::kable(result$summary, digits = 4,
             caption = "Top 3 non-hierarchical combinations")
```

## Next steps

| Goal | Vignette |
|------|----------|
| Hierarchical analysis, ICD-11 comparison, model validation | [Full Internal Analysis](internal_analysis.html) |
| Derive on one PCL-5 dataset, validate on another | [External Validation (PCL-5)](external_validation_pcl5.html) |
| Validate against the CAPS-5 gold standard | [External Validation (CAPS-5)](external_validation_caps5.html) |
| Full function reference | [Reference](../reference/index.html) |
